---
title: "Assignment: Using Sobol with an ODE"
author: "Lucian Scher"
date: "2/25/26"
format:
  pdf: default
  html: default
execute:
  echo: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(deSolve)
library(sensitivity)
```

This homework assignment models forest growth over 300 years using the
ODE solver and a sobol global sensitivity analysis.

> Model Outline from assignment outline:
Consider the following model of forest growth (where forest size in measured in units of carbon (C))
-   $dC/dt  = r*C$ for forests where C is below a threshold canopy closure
-   $dC/dt = g*(1- C/K)$ for forests where carbon is at or above the threshold canopy closure
-   and $K$ is a carrying capacity in units of carbon
The size of the forest ($C$), Canopy closure threshold and carrying capacity are all in units of carbon
You could think of the canopy closure threshold as the size of the forest at which growth rates change from exponential to linear You can think of $r$, as early exponential growth rate and $g$ as the linear growth rate once canopy closure has been reached

# 1: Implement differential equation model

```{r dforest_model}
#' Forest growth with canopy closure and carrying capacity
#'
#' @param Time time since start (years)
#' @param C forest carbon (kg C)
#' @param parms list with elements:
#'   - r: exponential growth rate before canopy closure
#'   - g: linear growth rate (kg C / year) after canopy closure
#'   - C_thresh: canopy closure threshold (kg C)
#'   - K: carrying capacity (kg C)
#'
#' @return list with derivative dC/dt
dforest_canopy <- function(Time, C, parms) {
  # unpack parameters
  r        <- parms$r
  g        <- parms$g
  C_thresh <- parms$C_thresh
  K        <- parms$K
  
  # choose equation based on whether forest is below or above canopy closure
  if (C < C_thresh) {
    # before canopy closure: exponential growth
    dC <- r * C
  } else {
    # after canopy closure: linear growth with carrying capacity
    dC <- g * (1 - C / K)
  }
  
  return(list(dC))
}
```

# 2: Run the Model

> Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kg/C, and using the following parameters:
-   canopy closure threshold of 50 kgC
-   $K$ = 250 kg C (carrying capacity)
-   $r$= 0.01 (exponential growth rate before before canopy closure)
-   $g$ = 2 kg/year (linear growth rate after canopy closure)

```{r run-base-model}
# initial conditions and time vector
Cinitial <- 10          # initial forest size (kg C)
simtimes <- seq(from = 0, to = 300, by = 1)  # years

# parameter list for the base case
base_parms <- list(
  r        = 0.01,
  g        = 2,
  C_thresh = 50,
  K        = 250
)

# run the ODE solver
base_result <- ode(
  y     = Cinitial,
  times = simtimes,
  func  = dforest_canopy,
  parms = base_parms
)

# name columns and convert to data frame
base_result <- as.data.frame(base_result)
colnames(base_result) <- c("time", "C")

head(base_result)
tail(base_result)
```

# 3: Graph the results

>  Graph the results. Here you are graphing the trajectory with the parameters as given (e.g no uncertainty)

```{r plot-base-trajectory}
ggplot(base_result, aes(time, C)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  labs(
    title = "Forest Carbon Trajectory (Deterministic Parameters)",
    x     = "Time (years)",
    y     = "Forest Carbon C (kg C)"
  ) +
  theme_minimal()
```

# 4: Run sobol global sensitivity analysis

> Run a sobol global (vary all parameters at the same time) sensitivity analysis that explores how the estimated **maximum forest size** (e.g maximum of $C$ 300 years, varies with these parameters
-   pre canopy closure growth rate ($r$)
-   post-canopy closure growth rate ($g$)
-   canopy closure threshold and carrying capacity($K$)
Assume that parameters are all normally distributed with means as given above and standard deviation of 10% of mean value

Create the Sobol sampling matrices (X1 and X2) for the four parameters, using normal distributions with means as given in the base case and standard deviations equal to 10% of the mean. Then ensure that any negative values (which don't make sense for these parameters) are set to zero.

```{r sobol-sampling}
set.seed(123)

# number of parameter samples in each Sobol matrix
np <- 1000

# base means
mean_r        <- 0.01
mean_g        <- 2
mean_C_thresh <- 50
mean_K        <- 250

# standard deviations = 10% of mean
sd_r        <- 0.10 * mean_r
sd_g        <- 0.10 * mean_g
sd_C_thresh <- 0.10 * mean_C_thresh
sd_K        <- 0.10 * mean_K

# first Sobol sample matrix X1
r1        <- rnorm(np, mean = mean_r,        sd = sd_r)
g1        <- rnorm(np, mean = mean_g,        sd = sd_g)
Cthresh1  <- rnorm(np, mean = mean_C_thresh, sd = sd_C_thresh)
K1        <- rnorm(np, mean = mean_K,        sd = sd_K)

X1 <- cbind.data.frame(
  r        = r1,
  g        = g1,
  C_thresh = Cthresh1,
  K        = K1
)

# second Sobol sample matrix X2
r2        <- rnorm(np, mean = mean_r,        sd = sd_r)
g2        <- rnorm(np, mean = mean_g,        sd = sd_g)
Cthresh2  <- rnorm(np, mean = mean_C_thresh, sd = sd_C_thresh)
K2        <- rnorm(np, mean = mean_K,        sd = sd_K)

X2 <- cbind.data.frame(
  r        = r2,
  g        = g2,
  C_thresh = Cthresh2,
  K        = K2
)

# ensure no negative values for parameters where they don't make sense
X1 <- X1 |> map_df(pmax, 0)
X2 <- X2 |> map_df(pmax, 0)

head(X1)
```

Create the Sobol objecet using the `sobolSalt` function, which sets up the design for Sobol sensitivity analysis. We will later “tell” this object with the model outputs to compute the indices.

```{r sobol-object}
sens_forest <- sobolSalt(
  model = NULL,
  X1    = X1,
  X2    = X2,
  nboot = 300
)

colnames(sens_forest$X) <- c("r", "g", "C_thresh", "K")
head(sens_forest$X)
```

Define a metric function (taking the ODE output and returning a list), 

```{r forest-metric-function}
# compute maximum C and C at a target time (100 years)
compute_forest_metrics <- function(result, time_target = 100) {
  # result: data frame with columns time and C
  
  # maximum forest carbon over the full simulation
  maxC <- max(result$C)
  
  # forest carbon at (or just after) the target time
  idx <- which(result$time >= time_target)[1]
  idx <- ifelse(is.na(idx), nrow(result), idx)
  C_target <- result$C[idx]
  
  return(list(maxC = maxC, C_target = C_target))
}
```

Wwrapper that runs the ODE and computes the metrics.

```{r forest-wrapper}
# wrapper function: runs ODE and computes metrics
forest_wrapper <- function(r, g, C_thresh, K,
                           Cinitial, simtimes,
                           odefunc, metricfunc,
                           metricpars) {
  # create parameter list for ODE
  parms <- list(
    r        = r,
    g        = g,
    C_thresh = C_thresh,
    K        = K
  )
  
  # run the ODE solver (Euler robust for simple models)
  result <- ode(
    y      = Cinitial,
    times  = simtimes,
    func   = odefunc,
    parms  = parms,
    method = "euler"
  )
  
  # clean up
  result <- as.data.frame(result)
  colnames(result) <- c("time", "C")
  
  # compute metrics
  metrics <- metricfunc(result, time_target = metricpars$time_target)
  return(metrics)
}

# quick test with the first parameter set
test_metrics <- forest_wrapper(
  r         = sens_forest$X[1, "r"],
  g         = sens_forest$X[1, "g"],
  C_thresh  = sens_forest$X[1, "C_thresh"],
  K         = sens_forest$X[1, "K"],
  Cinitial  = Cinitial,
  simtimes  = simtimes,
  odefunc   = dforest_canopy,
  metricfunc = compute_forest_metrics,
  metricpars = list(time_target = 100)
)

test_metrics
```

# 5: Run Sobol Workflow and Boxplot of Maximum Forest Size

> Graph the results of the sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (**S** and **T**).

```{r run-sobol-workflow}
# apply wrapper to all parameter sets using pmap 
allresults <- as.data.frame(sens_forest$X) |>
  pmap(
    forest_wrapper,
    Cinitial   = Cinitial,
    simtimes   = simtimes,
    odefunc    = dforest_canopy,
    metricfunc = compute_forest_metrics,
    metricpars = list(time_target = 100)
  )

# extract metrics into a data frame
allres <- allresults |>
  map_dfr(`[`, c("maxC", "C_target")) |>
  as.data.frame()

# rename for clarity: C_target is C at 100 years
colnames(allres) <- c("maxC", "C_100")

head(allres)
summary(allres)
```

```{r boxplot-maxC}
# boxplot of maximum forest size across parameter sets
ggplot(allres, aes(x = "Max C over 300 years", y = maxC)) +
  geom_boxplot(fill = "forestgreen", alpha = 0.7) +
  labs(
    title = "Distribution of Maximum Forest Carbon (Sobol Samples)",
    x     = "",
    y     = "Maximum Forest Carbon C (kg C)"
  ) +
  theme_minimal()
```

```{r sobol-maxC-indices}
# Sobol indices for maximum forest size
sens_forest_maxC <- sensitivity::tell(sens_forest, allres$maxC)

# label rows
rownames(sens_forest_maxC$S) <- c("r", "g", "C_thresh", "K")
rownames(sens_forest_maxC$T) <- c("r", "g", "C_thresh", "K")

# first-order indices S
sens_forest_maxC$S

# total-effect indices T
sens_forest_maxC$T
```

# Extra Credit: Sobol Indices for Forest Size After 100 Years

> Compute Sobol indices for a second metric: forest size after a 100 years

I re-use the same Sobol design but “tell” it with the (C) at 100
years.

```{r sobol-C100-indices}
# Sobol indices for forest carbon at 100 years
invisible(capture.output(sens_forest_C100 <- sensitivity::tell(sens_forest, allres$C_100)))

rownames(sens_forest_C100$S) <- c("r", "g", "C_thresh", "K")
rownames(sens_forest_C100$T) <- c("r", "g", "C_thresh", "K")

# first-order indices S (C at 100 years)
sens_forest_C100$S

# total-effect indices T (C at 100 years)
sens_forest_C100$T
```

```{r boxplot-C100}
# boxplot of forest size at 100 years
ggplot(allres, aes(x = "C at 100 years", y = C_100)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Forest Carbon After 100 Years (Sobol Samples)",
    x     = "",
    y     = "Forest Carbon C (kg C) at 100 years"
  ) +
  theme_minimal()
```

# 6: Discussion

> In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

Based on the Sobol indices for maximum forest size, the carrying capacity parameter (K) has the largest total-order index, meaning long-term forest carbon storage is most sensitive to factors that determine ultimate stand size such as site productivity and long-term climate constraints. Growth rate parameters (r) and (g), along with the canopy closure threshold, have stronger influence on short- to medium-term biomass like our forest size at 100 years, which reflects the quickness at which forests respond to climate and management conditions. Climate-change impacts that reduce (K) will likely limit forest carbon storage, while changes that speed up or slow early growth mainly affect how quickly forests approach their potential.
