---
title: "Assignment 5: Sobol Sensitivity "
subtitle: "Atmospheric Conductance Sensitivity Analysis for High Wind, Low Vegetation Setting"
author: "Lucian Scher"
date: 2/8/2025
format: 
  html:
    toc: true
    code-fold: show
editor: 
  markdown: 
    wrap: 72
---

*Assignment Part I*

The results of the Sobol sensitivity analysis in the Uranium in
Groundwater Model paper helps environmental problem solving by
identifying which input parameters influence human health risk from
uranium exposure in groundwater the most. They found that uranium
concentration (Cw) dominates ingestion and dermal exposure models, with
intake rate (IR) and exposure frequency (EF) as secondary drivers. This
helps us understand how groundwater monitoring programs focused on
uranium concentration measurements, while recognizing that variations in
intake rate may require targeted public health interventions such as
water consumption advisories or alternative water supplies. The Cw-IR
interaction for ingestion and Cw-F interaction for dermal exposure,
explain how risk cannot be understood through just single-parameter
assessments. The problem needs management approaches that address
contaminant levels and exposure behaviors. Additionally, The low
sensitivity of body weight and surface area suggest needing less precise
characterization in risk models, allowing managers to allocate limited
resources toward more influential variables. This quantitative
prioritization framework supports evidence-based decision-making by
directing monitoring efforts, intervention strategies, and regulatory
focus toward parameters that most substantially affect health outcomes,
ultimately enabling more effective and cost-efficient protection of
vulnerable populations in contaminated groundwater regions.

*Assignment Part 2*

# Introduction

This analysis performs a Sobol sensitivity analysis on the atmospheric
conductance model practiced in class but with conditions higher
windspeeds lower vegetation.

## Model Background

Atmospheric conductance ($C_{at}$) describes how easily water vapor
diffuses from vegetation surfaces into the air. The model equations are:

$$C_{at} = \frac{v_m}{6.25 \cdot \left[\ln\left(\frac{z_m-z_d}{z_0}\right)\right]^2}$$

where:

\- $z_d = k_d \cdot h$ (displacement height)

\- $z_0 = k_0 \cdot h$ (roughness length)

\- $z_m = h + 200$ cm (measurement height, 200 cm above vegetation)

\- $h$ = vegetation height (cm)

\- $v$ = windspeed (cm/s)

\- $k_d$ = 0.7 (default)

\- $k_0$ = 0.1 (default)

## Parameter Distributions

-   **Windspeed** ($v$): Normal distribution, $\mu = 300$ cm/s,
    $\sigma = 50$ cm/s
-   **Vegetation height** ($h$): Uniform distribution, 3.5 - 5.5 m
    (350 - 550 cm)
-   $k_d$: Normal distribution, $\mu = 0.7$, $\sigma = 0.007$
-   $k_0$: Normal distribution, $\mu = 0.1$, $\sigma = 0.001$

# Setup

```{r setup}
#| message: false
#| warning: false

# Load required libraries
library(sensitivity)
library(tidyverse)

# Set seed
set.seed(10)
```

# Define Atmospheric Conductance Function

```{r define-function}
#| label: catm-function

#' Calculate atmospheric conductance
#' 
#' @param height Vegetation height (cm)
#' @param v Windspeed (cm/s)
#' @param k_d Displacement height coefficient (default = 0.7)
#' @param k_0 Roughness length coefficient (default = 0.1)
#' @param z_m_offset Height above vegetation for windspeed measurement (cm, default = 200)
#' @return Atmospheric conductance (mm/s)

Catm <- function(height, v, k_d = 0.7, k_0 = 0.1, z_m_offset = 200) {
  # Calculate measurement height (200 cm above vegetation)
  z_m <- height + z_m_offset
  
  # Calculate displacement height and roughness length
  z_d <- k_d * height
  z_0 <- k_0 * height
  
  # Check that z_m > z_d (measurement height must be above displacement height)
  if (z_m <= z_d) {
    warning("Measurement height (z_m) must be greater than displacement height (z_d)")
    return(NA)
  }
  
  # Check that z_0 > 0 (roughness length must be positive)
  if (z_0 <= 0) {
    warning("Roughness length (z_0) must be positive")
    return(NA)
  }
  
  # Calculate atmospheric conductance
  C_at <- v / (6.25 * (log((z_m - z_d) / z_0))^2)
  
  return(C_at)
}
```

# Generate Parameter Sets Using Sobol

```{r generate-parameters}
#| label: sobol-parameters

np <- 1000 # Number of parameters

# First set of random samples from parameter distributions

k_0_set1 <- rnorm(n = np, mean = 0.1, sd = 0.001) 
k_d_set1 <- rnorm(n = np, mean = 0.7, sd = 0.007)
v_set1 <- rnorm(n = np, mean = 300, sd = 50)
height_set1 <- runif(n = np, min = 350, max = 550) # in cm

# Combine into first parameter matrix
X1 <- cbind.data.frame(
  k_0 = k_0_set1,
  k_d = k_d_set1,
  v = v_set1,
  height = height_set1
)

# Second set of random samples
k_0_set2 <- rnorm(n = np, mean = 0.1, sd = 0.001)
k_d_set2 <- rnorm(n = np, mean = 0.7, sd = 0.007)
v_set2 <- rnorm(n = np, mean = 300, sd = 50)
height_set2 <- runif(n = np, min = 350, max = 550)

# Combine into second parameter matrix
X2 <- cbind.data.frame(
  k_0 = k_0_set2,
  k_d = k_d_set2,
  v = v_set2,
  height = height_set2
)

# Create Sobol sensitivity analysis object with 100 bootstrap samples
sens_Catm_Sobol <- sobolSalt(model = NULL, X1, X2, nboot = 100)
```

# Run Model for All Parameter Sets

```{r run-model}
#| label: run-conductance-model

# Extract parameters from Sobol object
parms <- as.data.frame(sens_Catm_Sobol$X)

# Ensure column names match parameter names
colnames(parms) <- colnames(X1)

# Run atmospheric conductance model for each parameter set
res <- pmap_dbl(parms, Catm)

# Store estimations in Sobol sensitivity object
sens_Catm_Sobol <- sensitivity::tell(sens_Catm_Sobol, res, res.names = "ga")
```

# Compute Sobol Sensitivity Indices

```{r sobol-indices}
#| label: calculate-indices

# First-order sensitivity indices
# Add names to output 
row.names(sens_Catm_Sobol$S) <- colnames(parms)

# Total effect sensitivity indices
# Add names to output 
row.names(sens_Catm_Sobol$T) <- colnames(parms)

# Extract total effect indices with confidence intervals for ranking
total_effects <- sens_Catm_Sobol$T

# Create a data frame for easier manipulation and ranking
te_df <- data.frame(
  parameter = rownames(total_effects),
  total_effect = total_effects[, "original"],
  lower_ci = total_effects[, "min. c.i."],
  upper_ci = total_effects[, "max. c.i."]
)

# Sort by total effect
te_df <- te_df[order(te_df$total_effect, decreasing = TRUE), ]

# second highest total effect for plotting
second_most_sensitive <- te_df$parameter[2]
```

# Visualize Results

## Graph of Uncertainty in Response Variable

```{r plot-distribution}
#| label: plot-conductance-distribution
#| fig-height: 5
#| fig-width: 7

# Combine parameters and results for plotting
both <- cbind.data.frame(parms, gs = sens_Catm_Sobol$y)

# Plot histogram of conductance values
ggplot(both, aes(x = gs)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = mean(both$gs), 
             col = "red", 
             linetype = "dashed", 
             linewidth = 1) +
  labs(
    title = "Distribution of Atmospheric Conductance Estimates",
    subtitle = "Accounting for parameter uncertainty",
    x = "Conductance (mm/s)",
    y = "Frequency"
  ) +
  theme_minimal() +
  annotate("text", 
           x = mean(both$gs), 
           y = Inf, 
           label = paste("Mean =", round(mean(both$gs), 2), "mm/s"),
           vjust = 1.5,
           hjust = -0.1,
           color = "red",
           size = 3.5)
```

## Graph of Relationship Between Output and Windspeed

```{r plot-windspeed}
#| label: plot-vs-windspeed
#| fig-height: 6
#| fig-width: 8

# Plot conductance against windspeed (x-axis) colored by height

color_col <- both[[second_most_sensitive]]
ggplot(both, aes(x = v, y = gs, color = color_col)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_gradient(low = "blue", high = "red", name = second_most_sensitive) +
  labs(
    title = "Atmospheric Conductance vs. Windspeed",
    subtitle = paste("Colored by height (2nd most sensitive parameter)"),
    x = "Windspeed (cm/s)",
    y = "Conductance (mm/s)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

# Discussion

## Sobol Sensitivity Indices

```{r display-indices}
#| label: show-indices

# Format first-order indices
first_order_df <- as.data.frame(sens_Catm_Sobol$S)
first_order_df <- first_order_df[, c("original", "min. c.i.", "max. c.i.")]
colnames(first_order_df) <- c("First-Order Index", "Lower CI (95%)", "Upper CI (95%)")
first_order_df <- first_order_df[order(first_order_df$`First-Order Index`, decreasing = TRUE), ]

# Format total effect indices
total_effect_df <- as.data.frame(sens_Catm_Sobol$T)
total_effect_df <- total_effect_df[, c("original", "min. c.i.", "max. c.i.")]
colnames(total_effect_df) <- c("Total Effect Index", "Lower CI (95%)", "Upper CI (95%)")
total_effect_df <- total_effect_df[order(total_effect_df$`Total Effect Index`, decreasing = TRUE), ]

# Display tables
knitr::kable(first_order_df, 
             digits = 4, 
             caption = "First-Order Sensitivity Indices")

knitr::kable(total_effect_df, 
             digits = 4, 
             caption = "Total Effect Sensitivity Indices")
```

## Sensitivity to windspeed change

In our analysis windspeed sensitivity is greater compared to our class settings with lower windspeeds and taller vegetation. Windspeed remains the dominant parameter (highest total effect index) because it appears directly in the numerator of the conductance equation, creating a linear relationship. The higher variability also creates a wider parameter range, which lets windspeed explain more variance in the output. The higher mean windspeed (300 cm/s) means that the same relative uncertainty translates to larger variations in conductance. The shorter vegetation affects the denominator structure but does not diminish windspeed's importance, as the direct multiplicative relationship in the numerator ensures windspeed remains the primary driver of conductance variability.
